A Triple Origin for Twin Blue Stragglers in Compact Binaries

### 2018-12-02 apjl.tex
Add files via upload

\begin{abstract}
In this Letter, we propose a potential formation mechanism for twin blue stragglers in compact binaries that involves mass transfer from an evolved outer tertiary companion on to the inner binary via a circumbinary disk.  We apply our hypothetical scenario to the observed double blue straggler system Binary 7782 in the old open cluster NGC 188, and show that its observed properties are easily and even naturally reproduced within the context our proposed model.  Within the context of the hypothetical formation mechanism proposed here, the presented work predicts the following properties for the post-mass transfer double BS tertiary:  (1)  For the outer tertiary orbit, the orbital period should lie between 220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days, assuming initial masses for the inner binary components of m$_{\rm 1} =$ 1.1 M$_{\odot}$ and m$_{\rm 2} =$ 0.9 M$_{\odot}$ and an outer tertiary mass of m$_{\rm 3} =$ 1.4 M$_{\odot}$. (2)  Larger final WD masses, and hence core masses for the donor at the time of mass transfer, should correspond to larger final outer orbital periods for the tertiary. (3)  For the inner binary, the rotational axes of the BSs should be aligned with each other and the orbital plane of the outer tertiary WD. (4)  The BSs in the inner binary should have roughly equal masses, independent of their initial masses.  This predicts that the initially lower mass MS star should accrete the most, and should hence be polluted more significantly by any accreted material.  This could be observable in the surface layers of a radiative star (i.e., He, C and O if the donor is an RGB star, and/or s-process elements if the donor is an AGB star). (5) Twin BSs in compact binaries formed from the proposed mechanism should be more frequent in younger clusters with ages $\lesssim$ 4-6 Gyr, since the donor will have  radiative envelope.

@@

In this Letter, we propose a potential formation channel for Binary 7782, and compact double BS binaries in general, which involves mass transfer from an outer tertiary companion on to an inner MS-MS binary.  In section~\ref{dyn}, we constrain the range of present-day orbital parameters for a hypothetical outer tertiary companion, using a combination of dynamical and stellar evolution-based constraints.  In Section~\ref{sims} we present the numerical simulations used to study the mass transfer process in our hypothetical triple system, computed using the \texttt{AMUSE} software package, in order to study the evolution of the inner and outer orbital parameters during mass transfer.  We summarize and discuss the implications of our results for compact double BS binaries and, more generally, mass transfer in stellar triples in Section~\ref{discussion}.

@@

Consider a hierarchical triple system with component masses m$_{\rm 1}$ and m$_{\rm 2}$ for the inner binary, and mass m$_{\rm 3}$ for the outer tertiary companion.  The inner and outer binary orbital semi-major axes are denoted a$_{\rm in}$ and a$_{\rm out}$, respectively.  We assume circular orbits for both the inner and outer orbits.

We consider a scenario where the outer tertiary companion is filling its Roche lobe and is hence transferring mass to the inner binary.  Using conservation of angular momentum, we can equate the specific angular momentum of the accreted mass at the inner Lagrangian point of the donor star to the final specific angular momentum of the accretion stream at the circularization radius about the inner binary:
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - r_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer tertiary companion, a$_{\rm c}$ is the semi-major axis of the orbit about the inner binary corresponding to the circularization radius and v$_{\rm orb,c}$ is the orbital velocity at a$_{\rm c}$.  The distance from the center of mass corresponding to the outer tertiary companion defined by the Roche lobe is given by \citep{eggleton83}: 
\begin{equation}
\label{eqn:roche}
R_{\rm L} = \frac{0.49q^{2/3}}{0.6q^{2/3} + {\rm ln}(1 + q^{1/3})}a_{\rm out},
\end{equation}
where the mass ratio q is defined as q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$ m$_{\rm 2}$).  Combining Equation~\ref{eqn:roche} with Equation~\ref{eqn:specangmom1}, we can solve for the circularization radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we require that a$_{\rm in} <$ a$_{\rm c}$.

@@

We use smoothed-particle hydrodynamics (SPH) simulations to compute the time evolution of the mass transfer process.

@@

\item The BSs in the inner binary should have roughly equal masses, independent of their initial masses.  This is because it is the lowest mass object that typically accretes the fastest, since its orbital velocity and distance relative to the circumbinary disk is typically the lowest.  This quickly brings the mass ratio toward unity.  This predicts that the initially lower mass MS star should accrete the most, and should hence be polluted more significantly by any accreted material.  This could be observable in the surface layers of a radiative star.  If the donor is an RGB star, the accretor will be enriched in mostly carbon, oxygen and helium.  If the accretor is an AGB star, it will be enriched in mostly s-process elements.

@@

@@

@@

@@

@@

@@

>>>>

\begin{abstract}
In this Letter, we propose a potential formation mechanism for twin blue stragglers in compact binaries that involves mass transfer from an evolved outer tertiary companion on to the inner binary via a circumbinary disk.  We apply our hypothetical scenario to the observed double blue straggler system Binary 7782 in the old open cluster NGC 188, and show that its observed properties are easily and even naturally reproduced within the context our proposed model.  Within the context of the hypothetical formation mechanism proposed here, the presented work predicts the following properties for the post-mass transfer double BS tertiary:  (1)  For the outer tertiary orbit, the initial orbital period should lie between 220 days $\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days, assuming initial masses for the inner binary components of m$_{\rm 1} =$ 1.1 M$_{\odot}$ and m$_{\rm 2} =$ 0.9 M$_{\odot}$ and an outer tertiary mass of m$_{\rm 3} =$ 1.4 M$_{\odot}$. (2)  Larger final WD masses, and hence core masses for the donor at the time of mass transfer, should correspond to larger final outer orbital periods for the tertiary. (3)  For the inner binary, the rotational axes of the BSs should be aligned with each other and the orbital plane of the outer tertiary WD. (4)  The BSs in the inner binary should have roughly equal masses, independent of their initial masses.  This predicts that the initially lower mass MS star should accrete the most, and should hence be polluted more significantly by any accreted material.  This could be observable in the surface layers of a radiative star (i.e., He, C and O if the donor is an RGB star, and/or s-process elements if the donor is an AGB star). (5) Twin BSs in compact binaries formed from the proposed mechanism should be more frequent in younger clusters with ages $\lesssim$ 4-6 Gyr, since the donor will have  radiative envelope.

@@

In this Letter, we propose a potential formation channel for Binary 7782, and compact double BS binaries in general, which involves mass transfer from an outer tertiary companion on to an inner MS-MS binary.  In section~\ref{dyn}, we constrain the range of initial (i.e., pre-mass transfer) orbital parameters for a hypothetical outer tertiary companion, using a combination of dynamical and stellar evolution-based constraints.  In Section~\ref{sims} we present the numerical simulations used to study the mass transfer process in our hypothetical triple system, computed using the \texttt{AMUSE} software package, in order to study the evolution of the inner and outer orbital parameters during mass transfer.  We summarize and discuss the implications of our results for compact double BS binaries and, more generally, mass transfer in stellar triples in Section~\ref{discussion}.

@@

Consider a hierarchical triple system with component masses m$_{\rm 1}$ and m$_{\rm 2}$ for the inner binary, and mass m$_{\rm 3}$ for the outer tertiary companion.  The inner and outer binary orbital semi-major axes are denoted a$_{\rm in}$ and a$_{\rm out}$, respectively.  We assume circular orbits for both the inner and outer orbits, and co-planar triples only since this is what has been observed for low-mass tertiaries \citep[e.g.][]{moe18,tobin18}.

We consider a scenario where the outer tertiary companion is filling its Roche lobe and is hence transferring mass to the inner binary.  The mass transfer stream gathers at the circularization radius a$_{\rm c}$, and forms a circumbinary disk.  Using conservation of angular momentum, we can equate the specific angular momentum of the accreted mass at the inner Lagrangian point of the donor star to the final specific angular momentum of the accretion stream at the circularization radius about the inner binary:
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - r_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer tertiary companion, a$_{\rm c}$ is the semi-major axis of the orbit about the inner binary corresponding to the circularization radius and v$_{\rm orb,c}$ is the orbital velocity at a$_{\rm c}$.  The distance from the center of mass corresponding to the outer tertiary companion defined by the Roche lobe is given by \citep{eggleton83}: 
\begin{equation}
\label{eqn:roche}
R_{\rm L} = \frac{0.49q^{2/3}}{0.6q^{2/3} + {\rm ln}(1 + q^{1/3})}a_{\rm out},
\end{equation}
where the mass ratio q is defined as q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$ m$_{\rm 2}$).  Combining Equation~\ref{eqn:roche} with Equation~\ref{eqn:specangmom1}, we can solve for the circularization radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we require that a$_{\rm in} <$ a$_{\rm c}$.

@@

We use smoothed-particle hydrodynamics (SPH) simulations to compute the time evolution of the mass transfer process.  We adopt the same initial particle masses as in Figure~\ref{fig:fig1}, namely initial component masses of m$_{\rm 1}$ = 1.1 M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner binary components, and m$_{\rm 3}$ is computed for the outer tertiary according to our assumed mass ratio, with our fiducial case corresponding to q $=$ 0.7.  The initial orbital eccentricities are set to zero.  The inner and outer binary orbital semi-major axes are set to 0.1 AU and 1 AU, respectively (CORRECT).

@@

\item The BSs in the inner binary should have roughly equal masses, independent of their initial masses.  This is because it is the lowest mass object that typically accretes the fastest, since its orbital velocity and distance relative to the circumbinary disk is typically the lowest \citep[e.g.][]{haiman09,farris15,rafikov16,kelley17}.  This quickly brings the mass ratio toward unity.  This predicts that the initially lower mass MS star should accrete the most, and should hence be polluted more significantly by any accreted material.  This could be observable in the surface layers of a radiative star.  If the donor is an RGB star, the accretor will be enriched in mostly carbon, oxygen and helium.  If the accretor is an AGB star, it will be enriched in mostly s-process elements.

@@

\bibitem[\protect\citeauthoryear{Farris et al.}{2015}]{farris15} Farris B. D., Duffell P., MacFayden A. I., Haiman Z. 2014, ApJ, 783, 134 

@@

\bibitem[\protect\citeauthoryear{Haiman et al.}{2009}]{haiman09} Haiman Z., Kocsis B., Menou K. 2009, ApJ, 700, 1952  

@@

\bibitem[\protect\citeauthoryear{Kelley et al.}{2017}]{kelley17} Kelley L. Z., Blecha L., Hernquist L. 2017, MNRAS, 464, 3131 

@@

\bibitem[\protect\citeauthoryear{Moe \& Kratter}{2018}]{moe18} Moe M., Kratter K. M. 2018, ApJ, 854, 44 

@@

\bibitem[\protect\citeauthoryear{Rafikov}{2016}]{rafikov16} Rafikov R. R. 2016, ApJ, 827, 111 

@@

\bibitem[\protect\citeauthoryear{Tobin et al.}{2018}]{tobin18} Tobin J. J., Looney L. W., Li Z.-Y., Sadavoy S. I., Dunham M. M. Segura-Cox D., Kratter K., Chandler C. J., Melis C., Harris R. J., Perez L. 2018, ApJ, 867, 43 
